---
title: "Outlier Detection"
subtitle: "Using Mahalanobis Distance to find Outliers"
author: "Steffen Ruefer"
date: "18 December 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
```

## Introduction

A sample of the height-weight dataset is used to demonstrate outlier detection by using Mahalanobis 
distance. App link: <https://steffenruefer.shinyapps.io/mahalanobis_outliers/>

```{r, echo=FALSE}
df <- read.csv("data.csv")
df$Outlier <- "No"
cbPalette <- c("#999999","#E69F00", "#4288D5", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
ggplot(df, aes(x = weight, y = height, color = Outlier)) +
      geom_point(size = 4, alpha = 0.7) +
      labs(
            title = "Height vs Weight Plot",
            subtitle = "Detecting Outliers with Mahalanobis Distance",
            caption = "Source: http://wiki.stat.ucla.edu/socr/index.php/SOCR_Data_Dinov_020108_HeightsWeights"
      ) +
      xlab("Weight in kg") + ylab("Height in cm") +
      scale_x_continuous(breaks = seq(35, 80, 5)) +
      scale_y_continuous(breaks = seq(150, 200, 5)) +
      scale_colour_manual(values=cbPalette)
```


## Data Exploration

After loading the data, Mahalanobis Distance is calculated and a threshold is used to create a new
feature, classifying an observation as outlier (Yes or No).

```{r, echo=TRUE}
df <- read.csv("data.csv")
# Calculate Mahalanobis Distance and add to data frame
df$m_dist <- round(mahalanobis(df, colMeans(df), cov(df)),1)
df$Outlier <- "No"
df$Outlier[df$m_dist > 10] <- "Yes"
df$Outlier <- as.factor(df$Outlier)
head(df, 4)
```

## Server Calculations

In the server calculations, data for the plot and the output table are calculated every time the 
threshold is changed. Below shows part of the server calculation:

```{r, echo=TRUE, eval=FALSE}
shinyServer(function(input, output) {
      # Table Output - Create table of outliers
      output$table <- renderTable({
            df$Outlier <- "No"
            df$Outlier[df$m_dist > input$threshold] <- "Yes"
            df$Outlier <- as.factor(df$Outlier)
            # Table Output:
            dt <- df %>%
                  filter(Outlier == "Yes") %>%
                  select(height, weight, m_dist) %>%
                  arrange(desc(m_dist))
      })
})
```


## Resulting Plot

The Shiny Application will output a plot with outlier data as shown below:

```{r, echo=FALSE, eval=TRUE}
df <- read.csv("data.csv")
# Calculate Mahalanobis Distance and add to data frame
df$m_dist <- mahalanobis(df, colMeans(df), cov(df))
df$Outlier <- "No"
df$Outlier[df$m_dist > 20] <- "Yes"
df$Outlier <- as.factor(df$Outlier)

cbPalette <- c("#999999","#E69F00", "#4288D5", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
ggplot(df, aes(x = weight, y = height, color = Outlier)) +
      geom_point(size = 4, alpha = 0.7) +
      labs(
            title = "Height vs Weight Plot",
            subtitle = "Detecting Outliers with Mahalanobis Distance",
            caption = "Source: http://wiki.stat.ucla.edu/socr/index.php/SOCR_Data_Dinov_020108_HeightsWeights"
      ) +
      xlab("Weight in kg") + ylab("Height in cm") +
      scale_x_continuous(breaks = seq(35, 80, 5)) +
      scale_y_continuous(breaks = seq(150, 200, 5)) +
      scale_colour_manual(values=cbPalette)
```



